---
title: "My Test RNA-seq Analysis for GSE236155"
author: "B293443"
date: "24/03/2025"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## RNA-seq pipeline
-"mytable_feaures" - The alignment and annotation (feature counting) of the alignment
-You have QC data from FastQC and MultiQC- this can be saved for reference, but you do not need to bring this data into the R pipeline. If you really want you may refer to it from within R if there is a particular feature that you'd like to track through the analysis, but this is quite an unusual situation.

-I also make a small targets file in R, using some information obtained from GEO
- I cut out some of the analysis from the 'advanced pipeline' that was not generally relevant.
-Since the paper from my study presented a volcano plot with some candidates, I have created my own plot and labelled with their candidate genes to see how the analysis outcomes compare.
-I've also performed statistical analysis.  Actually, there is a slight issue with the version of DESeq2 currently running, so some features are not available- see the comments in the text where the model is created.


## Analysis of RNA-seq data using R 


```{r init, echo=F}
#Load the required libraries & load the files for the tutorial
library(Rsubread)
library(edgeR)
library(limma)
library(gplots)
library(DESeq2)
library(affy)
library(QuasR)
library(pheatmap)
library(EnhancedVolcano)
getwd()
load(file = "GSE236155/mytable_feaures")
#We now have files for the tutorial and also data loaded into the workspace- use ls() and dir() to view this data
ls()
dir()


```
Once we have loaded the data we are ready to run some further analysis.

```{r input_processing, echo=TRUE}
#Some further processing to access count data and column names.
table_rnaseq<-(mytable_feaures)$counts
colnames_rnaseq <-colnames(table_rnaseq)

#Here I want to generate an 'adf data frame' for the annotation.  
#This can be loaded from file but here I generate in  code for a change...
adf <-data.frame(Colnames =colnames_rnaseq)
library(stringr)
locc <- str_locate_all(pattern="Aligned.sortedByCoord.out.bam",adf$Colnames)
adf$ShortGEOName <-substr(adf$Colnames,1,locc[[1]][,"start"]-1)

#From the GEO page for this study- https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE236155
#GSM7518952	mouse_NC_1
#GSM7518953	mouse_NC_2
#GSM7518954	mouse_NC_3
#GSM7518955	mouse_OE_1
#GSM7518956	mouse_OE_2
#GSM7518957	mouse_OE_3

#My samples are in the same order as the annotaion so I can just create some short names and add to annotation
adf$ShortNames <-c("mouse_NC_1","mouse_NC_2","mouse_NC_3","mouse_OE_1","mouse_OE_2","mouse_OE_3")

#while I am doing this I will create a factor for the treatment
split_names <-str_split(adf$ShortNames, pattern="_")
treatment <-c()
for(i in 1:length(split_names)){
  treatment <-c(treatment,split_names[[i]][2])
}
adf$Treatment <- treatment

#Add some colours as well
#See https://stackoverflow.com/questions/42288075/how-to-autogenerate-color-vector-from-a-numeric-vector
z <- factor(adf$Treatment)
adf$Colours <-rainbow(length(levels(z)))[z]


#make sure the colnames of the study match the rownames of the annotation...
rownames(adf) <-adf$Colnames
if(identical(colnames_rnaseq,rownames(adf))==FALSE){
  print("Names do not match")
}else{
  print("Names match.. OK to proceed")
}

#OK switch to shortnames to help with the figures
#We can only swap this way because column order to consistent with adf
rownames(adf) <-adf$ShortNames
colnames(table_rnaseq) <-adf$ShortNames

colnames_rnaseq <-colnames(table_rnaseq)

```


## Loading the Data into DESeq
DESeq2 will used for the statistical analysis and the normalisation of the data. 
```{r loading_deseq, echo=TRUE}
#Create object for DESeq
#NOTE the design for differential analysis is fed into this object 
dds_rnaseq <- DESeqDataSetFromMatrix(countData = table_rnaseq,colData = adf,design = ~ Treatment)
#check the object dimensions
dim(dds_rnaseq)
head(rownames(dds_rnaseq))
colnames(dds_rnaseq)
#check the object type
class(dds_rnaseq)
typeof(dds_rnaseq)

#check what this object contains
slotNames(dds_rnaseq)
colnames(colData(dds_rnaseq))
colnames(assay(dds_rnaseq))

```

Quality Filtering Genes
Apply filters to remove genes that do not pass quality filters.


```{r apply_filters, echo=TRUE}
#first filter based upon counts
keep <- rowSums(counts(dds_rnaseq)) > 1
dds_rnaseq <- dds_rnaseq[keep,]
nrow(dds_rnaseq)
```

## Normalisation of Data

```{r normalisation, echo=TRUE}
#size factor estimation for normalisation- these are used to scale by library size when data is exported.
dds_rnaseq <- estimateSizeFactors(dds_rnaseq)
sizeFactors(dds_rnaseq)

```

## DESeq2 also provides methods that outputs data with various normalisations.
Here we generate multiple different normalisations and compare between then- but for the rest of the tutorial we use rlog normalisation!  Note that the normalised data is not used for differential expression analysis- just for data visualisation.


```{r plot_normalised, echo=T}

#Normalised log counts
counts_rnaseq <-log2(counts(dds_rnaseq, normalized=TRUE) +1)
#fpm is similar to cpm and discussed here https://rdrr.io/bioc/DESeq2/man/fpm.html.  Here the data is log transformed
fpm_rnaseq <-log(fpm(dds_rnaseq) +1)#this is also normalised by the sizeFactors
#rlog and vsd are discussed here https://genomebiology.biomedcentral.com/articles/10.1186/s13059-014-0550-8
vsd_rnaseq <- vst(dds_rnaseq, blind = T)
head(vsd_rnaseq, 3)
rld_rnaseq <- rlog(dds_rnaseq, blind = T)
head(rld_rnaseq, 3)


```
We will use rlog for the rest of this tutorial, but here we compare the different normalised data

##Plots with Normalised Data.

```{r build_plots, echo=TRUE}

#Boxplots need some labels..
boxplot(counts_rnaseq, main="RNA-seq Counts Data")
boxplot(fpm_rnaseq, main="RNA-seq Fragments per Million")
boxplot(assay(vsd_rnaseq), main="RNA-seq VSD (variance stabilized data)")#etc
boxplot(assay(rld_rnaseq), main="RNA-seq RLD (rlog transformed data)")#etc
#but suppose we look at non-normalised
na.rm=T
counts_un <-counts(dds_rnaseq, normalized=FALSE)+1

boxplot(log2(counts_un), main="RNA-seq unnormalized Counts Data")
counts_un_rlog<-rlog(counts_un)





#We can make an mva.pairs plot for any of these data
mva.pairs(counts_un_rlog,na.rm=TRUE)

```

Exploratory Data Plots
PCA plots similar to those we use for microarrays.


```{r more_explore, echo=T}
#we can run a PCA plot here...
library(scatterplot3d)
library(ggplot2)
# Perform PCA
pca <- prcomp(t(na.omit(assay(rld_rnaseq))), scale=T)
# Plot the PCA results
s3d<-scatterplot3d(pca$x[,1:3], pch=19, color=adf$Colours)
s3d.coords <- s3d$xyz.convert(pca$x[,1:3])
text(s3d.coords$x, s3d.coords$y, labels = colnames(rld_rnaseq),pos = 3,offset = 0.5,cex=0.5)
#or a quick plot in 2d
qplot(pca$x[,1],pca$x[,2],xlab="PCA1", ylab="PCA2",color=adf$Colours)
#We can also output a distance matrix…
sampleDists <- dist(t(assay(rld_rnaseq)))
sampleDists

```
## Generate further heatmaps with the data

```{r more_heatma-s, echo=T}
library("pheatmap")
library("RColorBrewer")
sampleDistMatrix <- as.matrix(sampleDists)
pheatmap(sampleDistMatrix)
#easier PCA plot
ress <-plotPCA(rld_rnaseq, intgroup = c("Treatment"))
ress <- ress + geom_label(aes(label = name))
plot(ress)
```


## Differential Gene Expression
Simplified stats

```{r differential_expression, echo=T}
dds_rnaseq <- DESeq(dds_rnaseq)

#We should apply this here https://support.bioconductor.org/p/77461/#107693
#But there is a bug in this version of DESeq2!!
#dds_rnaseq <- lfcShrink(dds_rnaseq, coef=2)




#This is how you specify a contrast...
#note: results(dds, contrast=c("condition","A","B")): genes with logFC > 0 overexpressed in A https://support.bioconductor.org/p/134510/
result_treatment <-results(dds_rnaseq,contrast=c("Treatment","OE","NC"))






#make a nice plot of differential expression
plotMA(result_treatment, main="DESeq2 dispersion plot", ylim=c(-2,2))
table(result_treatment$padj < 0.01)
table(result_treatment$padj < 0.05)
result_treatment_selected <- subset(result_treatment, padj < 0.05)
result_treatment <-result_treatment[order(abs(result_treatment$stat),decreasing = TRUE), ]
head(result_treatment)
```

## Make heatmaps from each of the output tables- selected top 50 or all using fold change and FDR.

```{r differential_expression_heatmaps, echo=T}
top70 <-rownames(result_treatment)[1:70]
#make a heatmap for the top 70
pheatmap(assay(rld_rnaseq)[top70,],scale="row",show_rownames=F,main="RLD values, Row Scaled, Top 70")
#all
pheatmap(assay(rld_rnaseq)[rownames(result_treatment),],scale="row",show_rownames=F,main="Treatment Selected, Row Scaled")


```


## Build Gene Annotation. A serialised is generated in case Ensembl is unavailable

```{r build_gene_annotation, echo=TRUE}
#Download Ensembl annotation using BiomaRt and rename the samples
library(biomaRt)
if(file.exists("resultAnnot.RData")){

  load("resultAnnot.RData")
  print("Loaded existing Ensembl annotation...")
}else{
  
  #UK ensembl is being updated  we use a USA mirror, "useast.ensembl.org"
  ensembl_host <-"https://www.ensembl.org"
  head(biomaRt::listMarts(host = ensembl_host), 15)
  head(biomaRt::listAttributes(biomaRt::useDataset(dataset = "mmusculus_gene_ensembl",mart = useMart("ENSEMBL_MART_ENSEMBL",host = ensembl_host))), 40)
  mart <- biomaRt::useDataset(dataset = "mmusculus_gene_ensembl",mart    = useMart("ENSEMBL_MART_ENSEMBL",host    = ensembl_host))
  resultAnnot <- biomaRt::getBM(values=rownames(dds_rnaseq),attributes =     c("ensembl_gene_id","external_gene_name","chromosome_name","start_position","end_position","description","strand"),filters="ensembl_gene_id",mart=mart)
  save(resultAnnot,file="resultAnnot.RData")
}

```

Merge Annotation with Input Data
```{r merge_gene_annotation, echo=TRUE}
#merge with input data
names <- resultAnnot[,1]
resultAnnot <- as.data.frame(resultAnnot)
resultAnnot <- resultAnnot[!duplicated(resultAnnot$ensembl_gene_id), ]
idx<-match(rownames(dds_rnaseq),resultAnnot$ensembl_gene_id)
resultAnnot <- resultAnnot[idx, ]

unmatched_ids <- rownames(dds_rnaseq)[is.na(resultAnnot$ensembl_gene_id)]
resultAnnot$ensembl_gene_id[is.na(resultAnnot$ensembl_gene_id)] <- unmatched_ids

all(resultAnnot$ensembl_gene_id == rownames(dds_rnaseq))
rownames(resultAnnot) <- resultAnnot$ensembl_gene_id
all(rownames(dds_rnaseq) == rownames(resultAnnot))

grr<-resultAnnot[match(rownames(dds_rnaseq), resultAnnot$ensembl_gene_id),]
all(rownames(dds_rnaseq) == rownames(grr))
resultAnnot <-grr
all(rownames(dds_rnaseq) == rownames(resultAnnot))

#make the nice names
names <- names[idx]
resultAnnot$ensembl_gene_id <- names
nice_names<-ifelse(resultAnnot$external_gene_name == resultAnnot$ensembl_gene_id, 
                   resultAnnot$ensembl_gene_id, 
                   paste(resultAnnot$ensembl_gene_id,resultAnnot$external_gene_name, sep = '_'))
resultAnnot$nice_names <-nice_names
head(resultAnnot)
all(rownames(dds_rnaseq) == rownames(resultAnnot))

#check names
rld_rnaseq <- rlog(dds_rnaseq, blind = TRUE)#
idx2 <-match(rownames(result_treatment)[1:50],rownames(dds_rnaseq))
plotme <-(rld_rnaseq)[rownames(result_treatment)[1:50],]

rownames(plotme)<-resultAnnot$nice_names[idx2]
#make heatmap with candidate genes
png(filename="heatmap_candidates.png",width=800, height=1000)
pheatmap(assay(plotme),scale="row",fontsize_row = 10,cellheight =12, cellwidth =12,treeheight_row = 40, treeheight_col = 40)
dev.off()

pheatmap(assay(plotme),scale="row",fontsize_row = 10,cellheight =12, cellwidth =12,treeheight_row = 80, treeheight_col = 100)

```
## Final Results Table
Since we have annotation that matches gene names to candidates we can also annotate the results table from differential
expression

```{r merge_gene_annotation2, echo=TRUE}

#this has the results ordered

#before 
head(result_treatment,50)

annot_2 <- resultAnnot
idx3 <- match(rownames(result_treatment), annot_2$ensembl_gene_id)

new_names <- annot_2$nice_names[idx3]
new_names[is.na(new_names)] <- rownames(result_treatment)[is.na(new_names)]

rownames(result_treatment)<-new_names
#after
print(result_treatment[1:50,])

#load into table viewer
#View(as.data.frame(result_treatment[1:50,]))

```

## The source paper showed candidates in a volcano plot- this is to compare

Current row names are not useful here- we need to use genes but then will discard any
without known names. This is OK here because we are exploring the annotation -genes without
a gene name and so are no use for annotating the data.

We can still keep unannoted genes as candidates!

```{r make_volcano, echo=TRUE}

result_treatment$gene<-annot_2$external_gene_name[idx3]
write.csv(as.data.frame(result_treatment), "result_treatment.csv")
top10 <- as.data.frame(result_treatment)
top10 <- top10[order(top10$stat), ]
top10 <- top10[1:10, ]
top10_idx <- match(rownames(top10), resultAnnot$external_gene_name)
top10$ensembl_gene_id <- resultAnnot$ensembl_gene_id[top10_idx]
write.csv(top10, "result_treatment_top10DEGs.csv")
has_annot <- is.na(result_treatment$gene)
has_annot <-!has_annot
result_treatment<-result_treatment[has_annot,]
rownames(result_treatment)<-result_treatment$gene

#Now we have gene names so we can plot the remaining genes in a volcano plot

#there were highlighted as interesting in the paper
paper_genes <-c("Cd36","Abca1","Abcg1","Abcg2","Abcg3","Abca8a","Abca8b","Abca9","Pparg")


#just check they're measured in this annotation
if(sum(paper_genes %in% rownames(result_treatment))==length(paper_genes)){
  print("all genes found in data")
}#you'd just include a subset to label if some genes were missing- here they are all present

#The paper found...
#for volvano see 
#https://bioconductor.org/packages/devel/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html
volcano_test <-EnhancedVolcano(result_treatment,
    lab = rownames(result_treatment),
    x = 'log2FoldChange',
    y = 'padj',
    selectLab = paper_genes,
    xlab = bquote(~Log[2]~ 'fold change'),
    ylab = bquote(~-Log[10] ~ "adjusted p-value"),
    subtitle = bquote(italic("p <0.05, |log2FC|≥1")),
    pCutoff = 0.05,
    FCcutoff = 1.0,
    max.overlaps = Inf,
    pointSize = 1.0,
    labSize = 1.5,
    labCol = 'black',
    labFace = 'bold',
    boxedLabels = TRUE,
    colAlpha = 4/5,
    legendPosition = 'right',
    legendLabSize = 10,
    legendIconSize = 4.0,
    drawConnectors = TRUE,
    widthConnectors = 1.0,
    colConnectors = 'black')

plot(volcano_test,xlim=800,ylim=800)
png("volcano_test.png",width = 800, height = 800)
plot(volcano_test)
dev.off()

#and look at the table
#View(as.data.frame(result_treatment[paper_genes,]))

print(as.data.frame(result_treatment[paper_genes,]))
write.csv(as.data.frame(result_treatment[paper_genes,]), "paper_genes.csv")


```

## Session Information
```{r session_info, echo=TRUE}

sessionInfo()

```


